YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending

### 2015-03-30 draft.Rnw
Inline fixes. Plots tweaked. Elaboration of feature selection result.

\documentclass[12 pt]{article}
\usepackage{graphicx, verbatim}
\usepackage{chronosys}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage[linktoc=all]{hyperref}
\usepackage{titlesec}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}
\begin{document}
\setcounter{tocdepth}{2}
\linespread{1.3}
\tableofcontents

<<init,echo=FALSE,warning=FALSE,results='hide',message=FALSE,cache=FALSE,eval=TRUE>>=
#DONT CACHE SAYS YIHUI
require(dplyr)              ### Used Extensively
require(ggplot2)            ### Used Extensively
require(gridExtra)          ### For facet plots
require(knitr)              ### For chunk control, compilation
require(leaps)              ### For feature selection
require(lubridate)          ### For date grouping, conversions
require(reshape2)           ### For melt() before ggplotx
require(scales)             ### For plots
require(texreg)             ### Regression Output using texreg()
require(tidyr)              ### For spread()
require(xtable)             ### Tabular Output using xtable()
options(scipen=999,digits=4,formatR.arrow=TRUE,width=90)
coruse = 'pairwise.complete.obs'
opts_chunk$set(list(fig.path='figure/minimal-', fig.align='center',echo=FALSE, fig.show='hold',tidy=TRUE,fig.height=5,fig.pos='H',warning=FALSE, message=FALSE,results='asis'))
@

@@

old = filter(nd, Date < dis1)
o1 = filter(old, Date <= dis0) %>% transform(grade=CreditGrade, regime="Auction-Early")
o2 =  filter(old, Date > dis0) %>% transform(grade=ProsperRating..Alpha., regime="Auction-Late")
old = rbind(o1,o2)
n = filter(nd, Date >= dis1) %>% transform(grade = ProsperRating..Alpha., regime="Prosper-rate")

@@

\title{The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\footnote{R code available at
\url{github.com/sshleifer/ProsperThesis}}\\ Senior Essay in Economics}
\date{Yale University, \today}
\maketitle 
\begin{abstract}
This essay analyzes the effect of information gathering costs on financial market
efficiency, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace.
Peer-to-peer lending is the practice of individual and institutional investors
lending money to borrowers over the internet without the assistance of a
traditional financial intermediary. Between 2006 and 2011, Prosper underwent two
major policy changes that guided and then forced investors to set higher
interest rates, and reduce their exposure to subprime borrowers.

During the policy regime before either policy intervention (2006-2008), investors set
interest rates by auction, made avoidably bad mistakes, and earned poor returns.
The data from the first regime contradicts three Efficient Markets
Hypothesis predictions: that (1) risky assets should earn positive expected returns,
(2) riskier assets should earn higher returns than less risky assets, and, most
conclusively, (3) ``prices should fully reflect publicly available information."

Controlling for macroeconomic differences between the time periods of each policy regime
reveals that contradictions (1) and (2), which disappeared after Prosper began
guiding investors towards setting higher interest rates, are partly the
consequence of a turbulent macroeconomic period, but that better investor
decisions, or Prosper setting interest rates from the beginning, would have
weakened the blow of the financial crisis. Contradiction (3), that strategies
using publicly available information can earn superior risk-adjusted returns,
which holds until Prosper begins directly setting rates, suggesting that 
centralized interest rate setting may lead to a more efficient market than the
unconstrained interaction of supply and demand, especially when information
gathering costs are high.
\end{abstract}
\newpage

\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama
((1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models such as CAPM imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets.\footcite{sharpe} The information aggregation perspective
associated with Hayek (1945) and Stiglitz and Grossman (1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

In the P2P Lending context, the market's price is the interest rate associated with a loan. This paper studies the question of whether interest rates on Prosper.com are set efficiently. (Fama, 1970)'s semi-strong EMH predicts that if interest rates were set efficiently, then...
\begin{enumerate}
\item Loans would have positive expected returns. Investors would be unwilling
  to lend to Borrowers if they expected to lose money. \footnote{Assuming a
  Risk-Free Rate of 0.}
\item Riskier loans would have higher expected returns. \footcite{sharpe}
\item It is not possible to do better than the average investor using publicly available information. \footcite{fama}
\end{enumerate}
I test hypothesis 1 by measuring the historical returns of loans up to a certain date, and then asking whether loans with negative expected returns get funded at that date.
I test hypothesis 2 by comparing the returns of high risk loans to low risk loans.
I test hypothesis 3 by using publicly available information to select loans with a higher than average expected return.

Prosper.com data provide some additional valuable insights into investor behavior. As it happens, over the period I study, there are three interest rate determination regimes:
\begin{enumerate}
\item Early Auction Regime (April 19, 2006-October 15, 2008): Rates are set by investor auction. Investors have large amounts of data, but Prosper does not guide them meaningfully.
\item Late Auction Regime: (July 1, 2009-December 20, 2010): Rates are set by investor auction, with Prosper offering valuable guidance in the form of more complex credit grading.
\item Prosper-rate Regime: (December 21, 2010 - August 3, 2011): Prosper sets rates using its more complex credit grading algorithm.
\end{enumerate}
Additionally, for regimes (2) and (3), subprime borrowers are not allowed to
apply for Prosper loans.

@@

In this spirit, I do not claim that poor investor decision making is the only
factor causing poor returns in the Early auction period, or that the impact of
the policy changes can be accurately isolated by my methodologies.
The contradictions with efficiency hypotheses 1 and 2 can certainly be partially
explained by the same bad luck that explains poor returns on all risky assets
during economically turbulent times.
I merely argue that returns in the early auction period were likely made worse
by poor investor decisions, as evidenced by the predictable mispricings that can
be exploited, as they are in Section 5, the core of my market inefficiency
argument, which finds that over a somewhat longer period of time, simple loan
selection strategies can use publicly available information to earn higher
returns than the benchmark.\footnote{Benchmark is calculated by assuming \$1 investment in every available loan.}

%if was all time, why would Prosper use same variables to make grades.
%Some of the evidence is inconsistent with time period argument. Can find variables, once Prosper starts using same variables, no more excess returns. Prosper themselves understood this.
\subsection{Background}
Peer-to-peer lending (``P2PL")  is the practice of lending money to borrowers
over the internet, without going through a traditional financial
intermediary.\footcite{p2pwiki} From April 2006 until December 2010, interest
rates were set by lenders\footnote{For the purposes of this essay, lender and
investor are synonyms.}, who competed for the lowest rate in a reverse auction
model.

In October 2008, Prosper was shut down by the SEC, and re-opened in July 2009
after a 9 month quiet period with more stringent under-writing standards, and
greater guidance for lenders. Importantly, all the information about borrower
creditworthiness that is used in this paper is similar to what was available to
investors when they bid on loans. Since December 2010, the beginning of the
\textit{Prosper-rate} regime, rates have been determined by Prosper, on the
basis of a borrower's credit history.

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. The minimum investment in a loan was \$25, with no
maximum, and lenders could hold as many loans as they wished.
Before a potential borrower's listing is posted on the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange, In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

\iffalse
\begin{table}
  \renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  \captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{5cm}}
    \toprule
    \addlinespace[1.5ex]
    1947 & AT and T Bell Labs develop the idea of cellular phones\\
    1968 & Xerox Palo Alto Research Centre envisage the 'Dynabook\\
    1971 & Busicom 'Handy-LE' Calculator\\
    1973 & First mobile handset invented by Martin Cooper\\
    1978 & Parker Bros. Merlin Computer Toy\\
    1981 & Osborne 1 Portable Computer\\
    1982 & Grid Compass 1100 Clamshell Laptop\\
    1983 & TRS-80 Model 100 Portable PC\\
    1984 & Psion Organiser Handheld Computer\\
    1991 & Psion Series 3 Minicomputer\\
  \end{tabular}
\end{table}
\fi
\iffalse
\startchronology
[startyear=2006,stopyear=2011,height=1ex,width=\hsize]
\chronoevent{0}{Jan 13 Quiz 1}
\chronoevent{8}{Jan 20 Quiz 2}
\chronoevent{36}{Feb 17 Quiz 6}
\chronoevent{36}{Intervention}
\stopchronology
\fi
In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of Loans funded on Prosper in each month of the sample. The second panel shows the total dollars lent to borrowers in each month of the sample.', fig.height=4>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup), Dollars = sum(LoanOriginalAmount))
g2 = melt(g2,id.vars = c("mgroup",'regime'))

ggplot(g2, aes(x=mgroup,fill=regime)) +
  geom_area(aes(y=value)) +
  facet_wrap(~variable,nrow=2,scale='free_y')+
  scale_x_date(labels=date_format('%b-%y')) +
  scale_y_continuous(labels=comma) +
  theme_bw()+
  labs(title='Prosper: Loan Origination by Month', x='Month', y='Number of Loans',fill='Regime',
       facet=c('Number of Loans','Dollars Lent'))
## TODO(fix facet labels)
@

<<catsTab, cache=FALSE>>=
catTab = filter(nd, purpose != "NA") %>% count(purpose, sort=TRUE) %>% mutate(Frequency = n/sum(n))
xtable(catTab, digits=3, label='catTab', caption = "Stated Reasons for Borrowing on Prosper (2006-2001). NAs removed")
@

@@

Second, whereas previously Prosper used only credit score to determine a borrower's Credit Grade (eg. AA), the site began offering a more advanced, multi variable formula to determine each borrower's credit grade for investors consideration. Finally, the site also partnered with FolioFN, a secondary market where lenders could liquidate their positions. Crucially, Prosper still had no direct role in setting interest rates, investors still used an auction, but with the extra information credit grades that more closely matched default probabilities. 
Besides credit score, Prosper began using (and still uses) the following ``Key variables" to determine its revamped credit grades: \footnote{CITE =  Estimated Loss Page}
\begin{itemize}
\item  TotalInquiries: The number of times potential lenders requested the borrowers' credit portfolio from the credit bureau, which is usually Experian or a competitor.
\item CurrentDelinquencies: The number of delinquent loans associated with the borrower.
\item Credit card utilization: how much of available credit the borrower uses on a monthly basis.
\item Number of open credit card accounts
\item Debt to income ratio
\item Loan payment performance on prior Prosper loans.
\end{itemize}
This Auction rate-setting mechanism continued for more than a year after the shutdown, until December 20, 2010.

Unsuprisingly the automated feature selection procedure in section 5, which
identifies the subset of variables that best predict default finds that
Credit Score, inquiries, delinquencies and number of credit cards are all useful
variables for predicting default.

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 21, 2010, lenders no longer determine the loan rate in an auction. Instead, Prosper uses the intelligent credit grades given to investors in the late auction regime to set interest rates on loans, with slightly less stringent under-writing standards to determine eligibility. The change was motivated by a desire to accelerate the listing process, so that investors could reinvest money more quickly. Moreover, towards the end of the late auction regime, auctions were arriving at interest rates that were almost completely determined by the credit score. Under the fixed-rate regime, investors choose whether or not to invest in a given loan, at the interest rate set by Prosper, so bids only consist of a dollar amount.\newline

\section{Literature}
\subsection{Efficient Markets Hypothesis}
In layman's terms, the efficient markets hypothesis (EMH), states that 'it is impossible to beat the market.' More specifically, Fama explains that market prices should always fully reflect publicly available information. Therefore, EMH rules out the possibility of strategies that earn excess returns using only publicly available information. \footcite{fama}
In the P2P lending context, this market price is the interest rate determined by
investors (or accepted by investors in the fixed-rate regime) at origination
time, and paid by the borrower, and returns are the cash flows received from borrowers divided by the amount invested.
(Fama, 1970) notes the dependence of most tests of market efficiency on a model of risk and return.
The models, including most famously, Sharpe's capital asset pricing model ('CAPM'), make two predictions:
First,  since an investor can always hold cash risklessly, the average return on risky investments must be positive. Second, securities with higher systematic risk earn higher average returns.\footcite{sharpe}

The EMH rests on three assumptions. ``First, investors are assumed to be rational and hence to value securities rationally. Second, to the extent that some investors are irrational, their decisions are random and uncorrelated. Third, to the extent that some investors are irrational in similar ways, they are met in the market by rational arbitrageurs, who eliminate their influence on prices." \footcite{as}

In a reverse interest rate auction, such as Prosper's market design, however, arbitrageurs cannot short a borrower and increase his interest rate.  In the situation where irrational investors under estimate a borrower's riskiness, they will outbid rational bidders, and set the market price. To earn excess returns, then, an investor must select loans for which other investors have demanded too high a rate, so that he can participate in funding.

\subsection{Soft Information and Credit Screening}
The novel idea of P2P is a large group of investors may be better equipped to assess risk than traditional financial institutions. ``Traditionally, the credit score provided by rating agencies has been the main tool banks use to screen smaller borrowers."\footcite{iyer} As such, the initial strand of literature focuses largely on the novel sources of information utilized by P2P investors.
(Gao and Lin, 2012) analyze the text written by the borrower in the ``description field" of the application, and find that descriptions that are easy to read have 2.3\% lower default rates than average, while complex narratives are associated with a 3.6\% higher default rate, but do not analyze return on investment.\footcite{gao} Pope and Sydnor (2008) find that investors can reduce their defaults by statistically discriminating against racial minorities, but do not analyze return on investment. (Iyer et. al,2009) find that within the credit score buckets provided by Prosper, investors use soft information to further sort borrowers within buckets, reducing defaults.\footcite{iyer}

These papers echo Hayek (1945)'s basic argument for market efficiency: that many
individuals, each using their own information, collectively arrive at one
efficient price.  The right price ``might have been arrived at by one single
mind possessing all the information which is in fact dispersed among all the
people involved in the process." \footcite{hayek}

Freedman and Jin (2008) are more focused on the failures of the auction regime,
Although the main focus of their paper is the role of soft information and
social networks on Prosper, they also study how investors learn to weight these
different variables, and confirm the finding of negative expected returns in the
early auction regime: ``lenders, especially those that joined
Prosper early, did not fully understand the market risk. We estimate, on
average, Lenders would have expected an internal rate of return of between
-.62\% and
1.38\% if they had correctly understood the risk distribution of Prosper
loans.''``Vigorous'' lender learning, they argue, explains the increase in interest
rates and borrower creditworthiness, and associated increase in lender
returns.\footcite{freedman}

@@

Rather differently than Hayek theorizes, the data suggest that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', that aggregates the data. Furthermore, as interest rate setting
becomes more centralized, interest rates are more sensitive to the hard credit variables
that actually predict default.
%t and other unused citations can be found in hypo.cut.tex

>>>>

\documentclass[12 pt]{article}
%\usepackage[doublespacing]{setspace}
\usepackage{graphicx, verbatim}
\usepackage[font=small,labelfont=sf]{caption}
\usepackage{chronosys}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage{titlesec}
\usepackage[linktoc=all]{hyperref}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}

\begin{document}
\setcounter{tocdepth}{2}
%\linespread{1.3}
\title{YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\footnote{R code available at
\url{github.com/sshleifer/ProsperThesis} TODO(THANKS(VK, Pope, AS, Nick, Iyer, Metrick)}}
\date{\today}
\maketitle
\begin{center}
Essay submitted to the Faculty of the Department of Economics in partial fulfillment
of the requirements for the degree of Bachelor of Arts at Yale University.

Thesis Advisor: Vineet Kumar\footnote{Yale School of Management}
\end{center}

\begin{center}
\textbf{Please print in color.}
\end{center}

\newpage
\begin{abstract}
This essay analyzes the effect of information gathering costs on financial market
efficiency, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace.
Peer-to-peer lending is the practice of individual and institutional investors
lending money to borrowers over the internet without the assistance of a
traditional financial intermediary. Between 2006 and 2011, Prosper underwent two
major policy changes that guided and then forced investors to lend at higher
interest rates, as well as to reduce their exposure to subprime borrowers.

During the policy regime before either policy intervention (2006-2008), investors set
interest rates by auction, made avoidably bad mistakes, and earned poor returns.
The data from this first regime is inconsistent with three of the predictions of
the Efficient Markets Hypothesis predictions: that (1) risky assets should earn positive expected returns,
(2) riskier assets should earn higher returns than less risky assets, and, most
conclusively, (3) ``prices should fully reflect publicly available information."

Controlling for macroeconomic differences between the time periods of each policy regime
reveals that inconsistencies (1) and (2), which disappeared after Prosper began
guiding investors towards setting higher interest rates, are partly the
consequence of a turbulent macroeconomic period. Inconsistency (3), that strategies
using publicly available information can earn superior risk-adjusted returns,
which holds until Prosper begins directly setting rates, suggests that 
centralized interest rate setting by a market maker may lead to a more efficient market than the
unconstrained interaction of supply and demand, especially when information
gathering costs are high.
\end{abstract}
\newpage
\tableofcontents

<<init,echo=FALSE,warning=FALSE,results='hide',message=FALSE,cache=FALSE,eval=TRUE>>=
#DONT CACHE SAYS YIHUI
require(dplyr)              ### Used Extensively
require(ggplot2)            ### Used Extensively
require(gridExtra)          ### For facet plots
require(knitr)              ### For chunk control, compilation
require(leaps)              ### For feature selection
require(lubridate)          ### For date grouping, conversions
require(reshape2)           ### For melt() before ggplot
require(scales)             ### For plots
require(texreg)             ### Regression Output using texreg()
require(tidyr)              ### For spread()
require(xtable)             ### Tabular Output using xtable()
options(scipen=999,digits=4,formatR.arrow=TRUE,width=90)
coruse = 'pairwise.complete.obs'
opts_chunk$set(list(fig.path='figure/minimal-', fig.align='center',echo=FALSE, fig.show='hold',tidy=TRUE,fig.height=5,fig.pos='H',warning=FALSE, message=FALSE,results='asis'))
@

@@

old = filter(nd, Date < dis1) #TODO (remove this)
o1 = filter(old, Date <= dis0) %>% transform(grade=CreditGrade, regime="Auction-Early")
o2 =  filter(old, Date > dis0) %>% transform(grade=ProsperRating..Alpha., regime="Auction-Late")
old = rbind(o1,o2)
n = filter(nd, Date >= dis1) %>% transform(grade = ProsperRating..Alpha., regime="Prosper-rate")

@@

\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama
(1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models such as CAPM imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets.\footcite{sharpe} The information aggregation perspective
associated with Hayek (1945) and Stiglitz and Grossman (1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

In the P2P Lending context, the market's price is the interest rate charged on a loan. This paper studies the question of whether interest rates on Prosper.com are set efficiently. (Fama, 1970)'s semi-strong EMH predicts that if interest rates were set efficiently, then...
\begin{enumerate}
\item Loans would have positive expected returns. Investors would be unwilling
  to lend to Borrowers if they expected to lose money. \footnote{Assuming a
  Risk-Free Rate of 0.}
\item Riskier loans would have higher expected returns. \footcite{sharpe}
\item It is not possible to do better than the average investor using publicly available information. \footcite{fama}
\end{enumerate}
Hypothesis 1 is tested by measuring historical returns of loans to borrowers of
each credit grade.
Hypothesis 2 is tested by comparing the returns of high risk loans to low risk
loans.
Hypothesis 3 is tested by using publicly available information to select loans
with a higher than average expected return, without look-ahead bias.

Prosper.com data provide some additional valuable insights into investor
behavior. Over the period I study, there are three interest rate determination regimes:
\begin{enumerate}
\item Early Auction Regime (April 19, 2006-October 15, 2008): Rates are set by investor auction. Investors have large amounts of data, but Prosper does not guide them meaningfully.
\item Late Auction Regime: (July 1, 2009-December 20, 2010): Rates are set by investor auction, with Prosper offering valuable guidance in the form of more complex credit grading.
\item Prosper-rate Regime: (December 21, 2010 - August 3, 2011): Prosper sets rates using its more complex credit grading algorithm.
\end{enumerate}
Additionally, for regimes (2) and (3), subprime borrowers are not allowed to
apply for Prosper loans.

@@

In this spirit, I do not claim that poor investor decision making is the only
factor causing poor returns in the Early auction period, or that the impact of
the policy changes can be fully isolated by my methodologies.
The inconsistencies with EMH predictions 1 and 2 can certainly be partially
explained by the same bad luck that explains poor returns on all risky assets
during economically turbulent times.
I merely argue that returns in the early auction period were likely made worse
by poor investor decisions, as evidenced by the predictable mispricings that
could be exploited. In Section 5, I show that over a somewhat longer period of time, simple loan
selection strategies can use publicly available information to earn higher
returns than the benchmark.\footnote{Benchmark is calculated by assuming \$1 investment in every available loan.}

Much of the evidence presented in this essay is inconsistent with an explanation
that all of the differences between regimes can be explained by differences in
time period. For example, in the late auction regime, when investors set
interest rates, strategies using publicly available information can earn excess
reutrns. In the Prosper-rate regime, when Prosper sets interest rates directly,
they cannot. The two periods are very similar macroeconomically. Perhaps most tellingly,
in the period when Prsper sets rates it seems to use precisely the information that was
helpful in selecting the best performing loans in the early auction regime.
The points strongly toward the conclusion that initial investors were not using
publicly available information.

\subsection{Background}
Peer-to-peer lending (``P2PL")  is the practice of lending money to borrowers
over the internet, without going through a traditional financial
intermediary.\footcite{p2pwiki} From April 2006 until December 2010, interest
rates were set by lenders\footnote{For the purposes of this essay, lender and
investor are synonyms.}, who competed for the lowest rate in a reverse auction
model.

In October 2008, Prosper was shut down by the SEC because it was designated as
a seller of securities,\footnote{Rather than, as it argued, a marketplace
connecting buyers and sellers.} and re-opened in July 2009
after a 9 month quiet period with more stringent under-writing standards, and
greater guidance for lenders. Importantly, all the information about borrower
creditworthiness that is used in this paper is similar to what was available to
investors when they bid on loans. Since December 2010, the beginning of the
\textit{Prosper-rate} regime, rates have been determined by Prosper, on the
basis of a borrower's credit history.

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. The minimum investment in a loan was \$25, with no
maximum, and lenders could hold as many loans as they wished.
Before a potential borrower's listing is posted on the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

\iffalse
\begin{table}
  \renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  \captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{5cm}}
    \toprule
    \addlinespace[1.5ex]
    1947 & AT and T Bell Labs develop the idea of cellular phones\\
    1968 & Xerox Palo Alto Research Centre envisage the 'Dynabook\\
    1971 & Busicom 'Handy-LE' Calculator\\
    1973 & First mobile handset invented by Martin Cooper\\
    1978 & Parker Bros. Merlin Computer Toy\\
    1981 & Osborne 1 Portable Computer\\
    1982 & Grid Compass 1100 Clamshell Laptop\\
    1983 & TRS-80 Model 100 Portable PC\\
    1984 & Psion Organiser Handheld Computer\\
    1991 & Psion Series 3 Minicomputer\\
  \end{tabular}
\end{table}
\fi
\iffalse
\startchronology
[startyear=2006,stopyear=2011,height=1ex,width=\hsize]
\chronoevent{0}{Jan 13 Quiz 1}
\chronoevent{8}{Jan 20 Quiz 2}
\chronoevent{36}{Feb 17 Quiz 6}
\chronoevent{36}{Intervention}
\stopchronology
\fi
In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of loans funded on Prosper.com in each month of the sample. The second panel shows the total dollars lent to borrowers on Propers.com in each month of the sample.'>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup) 
            ,Dollars_Lent = sum(LoanOriginalAmount)) %>%
    melt(id.vars = c("mgroup",'regime'))

ggplot(g2, aes(x=mgroup,fill=regime)) +
  geom_area(aes(y=value)) +
  facet_wrap(~variable,nrow=2,scale='free_y')+
  scale_x_date(labels=date_format('%b-%y')) +
  scale_y_continuous(labels=comma) +
  theme_bw()+
  labs(title='Prosper: Loan Origination by Month', x='Month', y='Number of Loans',fill='Regime')
@

<<catsTab, cache=FALSE>>=
catTab = filter(nd, purpose != "NA") %>% count(purpose, sort=TRUE) %>% mutate(Frequency = n/sum(n))
colnames(catTab) = c('Purpose','n','Frequency')
xtable(catTab, digits=3, label='catTab', caption = "Stated reasons for borrowing on Prosper (2006-2001). NAs removed.")
@

@@

Second, whereas previously Prosper used only the credit score to determine a borrower's Credit Grade (eg. AA), the site began offering a more advanced, multi variable formula to determine each borrower's credit grade for investors consideration. Finally, the site also partnered with FolioFN, a secondary market where lenders could liquidate their positions. Crucially, Prosper still had no direct role in setting interest rates, investors still used an auction, but with the extra information credit grades that more closely matched default probabilities. 
Besides credit score, Prosper began using (and still uses) the following ``Key
variables" to determine its revamped credit grades:\footcite{estloss}
\begin{itemize}
\item  TotalInquiries: The number of times potential lenders requested the borrowers' credit portfolio from the credit bureau, which is usually Experian or a competitor.
\item CurrentDelinquencies: The number of delinquent loans associated with the borrower.
\item Credit card utilization: how much of available credit the borrower uses on a monthly basis.
\item Number of open credit card accounts
\item Debt to income ratio
\item Loan payment performance on prior Prosper loans.
\end{itemize}
This Auction rate-setting mechanism continued for more than a year after the shutdown, until December 20, 2010.

Unsuprisingly, buut critically for the interpretation of the evidence,
the automated feature selection procedure in section 5, which
identifies the subset of variables that best predict default, finds that
credit Score, inquiries, delinquencies and number of credit cards are all useful
independent variables for predicting default.

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 21, 2010, lenders
no longer determine the loan rate in an auction. Instead, Prosper uses the
intelligent credit grades given to investors in the late auction regime to set
interest rates on loans, with slightly less stringent under-writing standards to
determine eligibility. The change was motivated by a desire to accelerate the
listing process, so that investors could reinvest money more quickly. Moreover,
towards the end of the late auction regime, auctions were arriving at interest
rates that were almost completely determined by the credit grade. Under the
fixed-rate regime, investors choose whether or not to invest in a given loan, at
the interest rate set by Prosper, so bids consist only of a dollar amount.\newline
%TODO: NEW PAGE IF CRUNCHED
\section{Literature}
\subsection{Efficient Markets Hypothesis}
In layman's terms, the efficient markets hypothesis (EMH), states that 'it is
impossible to beat the market.' More specifically, Fama explains that market
investors (or accepted by investors in the fixed-rate regime) at origination
time, and paid by the borrower, and returns are the cash flows received from borrowers divided by the amount invested.
Fama, (1970) notes the dependence of most tests of market efficiency on a model of risk and return.
The models, including most famously, Sharpe's capital asset pricing model ('CAPM'), make two predictions:
First,  since an investor can always hold cash risklessly, the average return on risky investments must be positive. Second, securities with higher systematic risk earn higher average returns.\footcite{sharpe}

The EMH rests on three assumptions. ``First, investors are assumed to be rational and hence to value securities rationally. Second, to the extent that some investors are irrational, their decisions are random and uncorrelated. Third, to the extent that some investors are irrational in similar ways, they are met in the market by rational arbitrageurs, who eliminate their influence on prices."\footcite{as}

In a reverse interest rate auction, such as Prosper's market design, arbitrageurs cannot short a borrower and increase his interest rate.  In the situation where irrational investors under estimate a borrower's riskiness, they will outbid rational bidders, and set the market price. To earn excess returns, then, an investor must select loans for which other investors have demanded too high a rate, so that he can participate in funding.

\subsection{Soft Information and Credit Screening}
The novel idea of P2P is a large group of investors may be better equipped to assess risk than traditional financial institutions. ``Traditionally, the credit score provided by rating agencies has been the main tool banks use to screen smaller borrowers."\footcite{iyer} As such, the initial strand of literature focuses largely on the novel sources of information utilized by P2P investors.
Gao and Lin (2012) analyze the text written by the borrower in the ``description
field" of the application, and find that descriptions that are easy to read have
2.3\% lower default rates than average, while complex narratives are associated
with a 3.6\% higher default rate, but do not analyze return on
investment.\footcite{gao} Pope and Sydnor (2008) find that investors can reduce
their defaults by statistically discriminating against racial minorities, but do
not analyze return on investment. Iyer et. al (2009) find that within the credit score buckets provided by Prosper, investors use soft information to further sort borrowers within buckets, reducing defaults.\footcite{iyer}

These papers echo Hayek (1945)'s basic argument for market efficiency: that many
individuals, each using their own information, collectively arrive at one
efficient price.  The right price ``might have been arrived at by one single
mind possessing all the information which is in fact dispersed among all the
people involved in the process."\footcite{hayek}

Freedman and Jin (2008) are more focused on the failures of the auction regime,
Although the main focus of their paper is the role of soft information and
social networks on Prosper, they also study how investors learn to weight these
different variables, and, like this paper, find negative expected returns in the
early auction regime: ``lenders, especially those that joined
Prosper early, did not fully understand the market risk. We estimate, on
average, Lenders would have expected an internal rate of return of between
-.62\% and
1.38\% if they had correctly understood the risk distribution of Prosper
loans.''``Vigorous'' lender learning, they argue, explains the increase in interest
rates and borrower creditworthiness, and associated increase in lender
returns.\footcite{freedman}

@@

Inconsistent with Hayek, the data suggest that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', namely a market maker who aggregates the data. Furthermore, as interest rate setting
becomes more centralized, interest rates are more sensitive to the hard credit variables
that actually predict default.
